---
title       : Monte-Carlo Simulation
author      : Dhananjay Ghei
date        : 17th Sep 2017
job         : UMN
output      :
            html_document:
                toc: true
                toc_float: true
                mathjax: local
                self_contained: false
geometry    : margin=0.5in
theme       : cerulean
---


<style type="text/css">

body{ /* Normal  */
   font-size: 14px;
}

h1 { /* Header 1 */
 font-size: 22px;
 color: DarkBlue;
}
h2 { /* Header 2 */
   font-size: 19px;
   color: DarkBlue;
}
td{font-size: 11pt;
   padding:0px;
   cellpadding="0";
   cellspacing="0"
}
th {font-size: 11pt;
   height: 20px;
   font-weight: bold;
   text-align: center;
   background-color:
   #ccccff;
}
table{border-spacing:
	0px;
	border-collapse:
	collapse;
}

</style>
								
# Monte-Carlo Simulation

This article demonstrates a method to do Monte-Carlo simulations in R
for a two variable linear regression. The source code for this exercise
is available [online](www.github.com/dhananjayghei/econometrics) in the
Github repository.

We will operate with the following model:

   $y_i = \beta_0 + x_i \beta_1 + \epsilon_i$
   
where, $y_i$ and $x_i$ are random variables. We observe $N$ samples
   of $i=1,2, \dots, n$ obersvations of $y$ and $x$.

We will estimate
   
   $y_i = \hat \beta_0 + x_i \hat \beta_1 + e_i$

```{r setOptions, message=FALSE, echo=FALSE, results='hide', warning=FALSE, include = FALSE}
source("monte_carlo.R")
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=6, echo=FALSE, warning=FALSE, message=FALSE)
```
# Question 1
First assume that the $\epsilon$ and the regressors are
uncorrelated. Assume that both $X$ and $\epsilon$ are drawn from a
normal distribution $\epsilon \sim N(0, \sigma_{\epsilon}^2), X \sim
N(\mu_x, \sigma_x^2)$. Start with a sample size ($n$) of
$n=100$ and define the number of samples to be simulated ($N$) to be
$N=100$ as well. You also need to define the true values of $\beta_0$
and $\beta_1$. Report the results that you fnid for the means and
variance-covariance matrix of $\hat \beta_0$ and $\hat
\beta_1$. Finally, plot the distribution of the estimated parameters
and the theoretical normal distribution that they are supposed to
follow. 

## Generating multivariate random variables $\epsilon$ and $X$


```{r geoDistribution,echo = FALSE, results = 'asis', strip.white = TRUE}
# Intercept (\hat \beta_0)
ggplot(bhat.Baseline, aes(x=b0hat))+geom_density(colour="midnightblue",
                                        fill="midnightblue", alpha=0.3) +
    theme_bw() + labs(x="", y="Density")+
    theme(legend.background=element_rect())

# Slope (\hat \beta_1)
ggplot(bhat.Baseline, aes(x=b1hat))+geom_density(colour="midnightblue",
                                        fill="midnightblue", alpha=0.3) +
    theme_bw() + labs(x="", y="Density") +
    theme(legend.background=element_rect()) 
```

The graph indicates that a significant percentage (22%) of the

# Question 2
Now vary the sample size ($n$) and redo 1) under the new
values. Report and plot your results. Comment.
```{r q2, echo = FALSE, results = 'asis', strip.white=TRUE}
ggplot(bhat.n, aes(x=b0hat, fill=type)) + 
    geom_density(alpha=0.5) +           
    theme_bw() +            
    labs(x="", y="Density") +
    theme(legend.background=element_rect())  +
    scale_fill_manual(values=c("midnightblue", "darkorange4"),
                      name = "",
                      breaks=c("Small Sample", "Large Sample"),
                      labels=c("Sample size - 10", "Sample size - 1000"))
```

# Question 3
Now vary the number of samples ($N$) and redo 1) under the new value.s
Report and plot your results. Comment.

```{r q3, echo = FALSE, results = 'asis', strip.white = TRUE}
ggplot(bhat.N, aes(x=b0hat, fill=type)) +
    geom_density(alpha=0.5) +
    theme_bw() +
    labs(x="", y="Density") +
    theme(legend.background=element_rect())  +
    scale_fill_manual(values=c("midnightblue", "darkorange4"),
                      name = "",
                      breaks=c("Small Sample", "Large Sample"),
                      labels=c("No. of samples - 10", "No. of samples - 1000"))
```

# Question 4
Now vary the distribution of X and redo 1) under the new
assumptions. Report and plot your results. Comment.

# Question 5
Now set the mean of the error $\epsilon$ to be different from zero and
redo 1) under the new values. Comment.

# Question 6
Now draw the $\epsilon$ in such a way that it is in fact correlated
with $X$ and redo 1) under the new assumptions. Report and plot your
results. Comment.

